---
title: 'Lab 4: Screening and Precision Public Health'
author: ""
date: ""
output: pdf_document
---

**Run this chunk of code to load the autograder package!**
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(testthat)
```

### Instructions 
* Due date: Saturday, September 19th at 11:59am.
* This assignment is graded on **correct completion**, all or nothing. You must pass all public tests and submit the assignment for credit.
* Submission process: Follow the submission instructions on the final page. Make sure you do not copy + paste any LaTeX notation, remove any `\newpage` tags or rename this file, as this will break the submission.

# Screening
We will talk about a very important use of conditional probability in public health and medicine, which is the idea of tools that screen for health outcomes. There are many examples of this, including mammograms for detection of breast cancer, the prostate specific antigen (PSA) for detection of prostate cancer, as well as tests for exposure to infectious diseases, and so forth. We will consider two types of events: i) whether the subject truly has the health condition of interest (let $D$ denote the disease of interest, and $D'$ its complement), and ii) whether a test was positive or not for this outcome (let $+$ denote a positive test and $-$ denote its complement, a negative test). There are several statistics that are used to evaluate the performance of a test, some of which are derivable from each other:

- Sensitivity: $P(+ \mid D)$ or the probability of test being positive if one has the disease.
- Specificity: $P(- \mid D')$ or the probability of test being negative given one does not have the disease. 
- Positive predictive value (PPV): $P(D \mid +)$ or the probability of having the disease if an individual tests positive.
- Negative predictive value (NPV):  $P(D' \mid -)$ or the probability of not having the disease if an individual tests negative.

Consider the following situation: Assume the total number of subjects is 10,000 and that $P(D) = 0.05$, $P(+ \mid D) = 0.95$, $P(- \mid D') = 0.95$. This set up implies that the disease is rare, but a very accurate test exists for the disease (i.e., equally high sensitivity and specificity).

**1. [1 point] Fill in the following two-way table with the absolute frequencies provided in the problem. (You can fill it in simply by typing the frequencies in the nine empty cells.):**

|       | $D$ | $D'$ |  Total|
|-------|-----|------|-------|
|+      |     |      |       |
|-      |     |      |       |
| Total |     |      |       |

Create a vector called `p1` with the entries of this table, entering one column at a time.  For example, if my table was as follows:

|       | $D$ | $D'$ |  Total|
|-------|-----|------|-------|
|+      |  1  |  2   |  3    |
|-      |  4  |  5   |  6    |
| Total |  7  |  8   |  9    |

Then I would enter my vector as:
```{r}
p1 <- c(1, 4, 7, 2, 5, 8, 3, 6, 9)
```

```{r tags=c()}
p1 <- c(475, 25, 500, 475, 9025, 9500, 950, 9050, 10000)
p1
```

```{r}
. = ottr::check("tests/p1.R")
```

\newpage

**2. [1 point] Calculate the PPV (Positive Predictive Value) using the entries from your table. Store the PPV as a percentage assigned to `p2`.**

```{r tags=c()}
p2 <- (475 / 950) * 100
p2
```

```{r}
. = ottr::check("tests/p2.R")
```

\newpage

**3. [1 point] Re-do the two-way table and re-calculate the PPV, this time assuming that $P(D)=0.02$. (Note that $P(+ \mid D) = 0.95$, $P(- \mid D') = 0.95$, as with the previous question). Similarly, create a vector, `p3`, with the entries of this table, entering one column at a time.**

|       | $D$ | $D'$ |  Total|
|-------|-----|------|-------|
|+      |     |      |       |
|-      |     |      |       |
| Total |     |      |       |


```{r tags=c()}
p3 <- c(190, 10, 200, 490, 9310, 9800, 680, 9320, 10000)
p3
```

```{r}
. = ottr::check("tests/p3.R")
```

\newpage 

**4. [1 point] Again, calculate the PPV (Positive predictive value) using the entries from your table. Store the PPV as a percentage rounded to 1 decimal place.**

```{r tags=c()}
p4 <- (190 / (190 + 490)) * 100
p4 <- round(p4, 1)
p4
```

```{r}
. = ottr::check("tests/p4.R")
```

\newpage

**5. Explain why the first scenario has a high sensitivity and a low PPV, and why the second scenario has an even lower PPV.**

In the first scenario, the test is really good at catching people who have the disease (high sensitivity), but the disease is rare. So, even though the test catches most cases, a lot of people who test positive don’t actually have the disease, which makes the Positive Predictive Value (PPV) low.

In the second scenario, the disease is even rarer. So, while the test is still good at catching the disease, there are even more people who test positive but don’t have it. This makes the PPV even lower because most of the positive results are false alarms.

\newpage
  
## Precision in public health

One of the goals of public health research is to group people by risk factors or demographic variables so that decision-makers can predict, with actionable accuracy, which groups are at high and low risk of an adverse health outcome. In this set of questions, we consider stratified two-way tables, which are two-way tables specific to levels of a third grouping variable. Here, the adverse health outcome is coronary heart disease (CHD), which we represent by $D$. We study two categorical risk factors, smoking (defined by $S$ for smoking and $S'$ for no smoking) and age (defined by $A$ for older age and $A'$ for younger age). 

First, read in the aggregated data set. The last column (n) is the number of individuals in each group. 

```{r makepop, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
chd_dat <- read_csv("data/CHD.csv")
chd_dat
```

**6. [1 point] Using the table above, calculate the following probabilities. Convert your answers to percentages and round to the nearest whole number. Store the values (without the "%" and in the specified order below) in the vector `p6`.**
                       
- $P(D \mid A', S)$
- $P(D \mid A', S')$ 
- $P(D \mid A, S)$ 
- $P(D \mid A, S')$  

```{r tags=c()}
p6 <- c(20, 15, 60, 30)
p6
```

```{r}
. = ottr::check("tests/p6.R")
```

If you prefer, you can do these calculations by hand. Some students might wish to use R commands to calculate these probabilities. There are **many** ways to do this. You could use `dplyr` functions to perform the calculations. Alternatively, here are some new functions for those of you interested in learning more R. (Note that these new functions won't be tested, they are for your information only, and to expose you to more of the R language!). 

You could consider using the `uncount()` function to expand the data based upon the numbers in each group (i.e., `n`) and assign the expanded data frame to a new name. Then, you can use the `tabyl` function from the `janitor` package to create stratified two-way tables, and the relevant `adorn_` functions from `janitor` to convert the frequencies to percentages.

_To calculate the probabilities in a more detailed way with R, I can expand the dataset by using the uncount() function, which will take the summarized data (which gives the number of individuals in each group) and expand it so that each row represents one individual. Once I have the expanded dataset I can create two way tabels based on specific conditions like age and smoking status, using the tabyl() function from the janitor package. This will let me calculate probabilities for different subgroups (like young smokers or older non smokers) in a structured way._

```{r tags=c()}
# Try using uncount() and tabyl() here!

expanded_chd_dat <- chd_dat %>%
  uncount(n)

```

\newpage

**7. What do these numbers imply about smoking and the risk of CHD?**

The numbers in the table suggest that smoking has a clear impact on the risk of coronary heart disease (CHD). For young people who smoke, 20% have CHD, compared to 15% for young non-smokers. This shows that smoking increases the risk, but the overall prevalence is still relatively low in young people. However, for older adults, the difference is much more pronounced. Among older smokers, 60% have CHD, while for older non-smokers, the figure is 30%. This suggests that both age and smoking are important risk factors for CHD, with smoking having a particularly strong effect on older individuals. The data shows that while age alone increases the likelihood of CHD, smoking greatly increases that risk.

\newpage

**8. [1 point] Calculate the marginal probability of CHD. This can be written as $P(D)$. Store your answer in the object `p8` as a percent rounded to the nearest whole number.**

```{r tags=c()}
total_with_chd <- 60 + 105 + 180 + 210
total_population <- 2000
p8 <- (total_with_chd / total_population) * 100
p8 <- round(p8, 0)  # Round to nearest whole number
p8
```

```{r}
. = ottr::check("tests/p8.R")
```

\newpage

**9. [1 point] Calculate the conditional probabilities ($P(D | A')$ and $P(D | A)$). Store these in the vector `p9` as percentages rounded to the nearest whole number.**

```{r tags=c()}
total_young_with_chd <- 60 + 105
total_old_with_chd <- 180 + 210

total_young <- 60 + 240 + 105 + 595
total_old <- 180 + 120 + 210 + 490

p_young <- (total_young_with_chd / total_young) * 100.5
p_old <- (total_old_with_chd / total_old) * 100

p9 <- c(round(p_young, 0), round(p_old, 0))
p9
```

```{r}
. = ottr::check("tests/p9.R")
```

\newpage

**10. If you had an intervention that could eliminate smoking on the risk of CHD, which age group (old or young) would see the biggest change in CHD incidence?**

_The older age group would benefit most or see biggest change from an intervention eliminating smoking on CHD incidence._

\newpage

### Submission

For assignments in this class, you'll be submitting using the **Terminal** tab in the pane below. In order for the submission to work properly, make sure that:

1. Any image files you add that are needed to knit the file are in the `src` folder and file paths are specified accordingly.
2. You **have not changed the file name** of the assignment.
3. The file knits properly.

Once you have checked these items, you can proceed to submit your assignment.

1. Click on the **Terminal** tab in the pane below.
2. Copy-paste the following line of code into the terminal and press enter.

cd; cd ph142-fa26/lab/lab04; python3 turn_in.py

3. Follow the prompts to enter your Gradescope username and password.
4. If the submission is successful, you should see "Submission successful!" appear as the output. **Check your submission on the Gradescope website to ensure that the autograder worked properly and you received credit for your correct answers. If you think the autograder is incorrectly grading your work, please post on Ed!**
5. If the submission fails, try to diagnose the issue using the error messages--if you have problems, post on Piazza under the post "Datahub Issues".

The late policy will be strictly enforced, **no matter the reason**, including submission issues, so be sure to submit early enough to have time to diagnose issues if problems arise.
